---
title: "Text and Sentiment Analysis"
description: |
  Text analysis of cowboy songs
theme: aj_theme.css
author:
  - name: AJ Zekanoski
date: 2021-03-13
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidytext)
library(textdata)
library(pdftools)
library(ggwordcloud)
```

Read in text of *Songs of the Cattle Trail and Cow Camp*.
```{r, cache = TRUE}
cowboys_text <- pdf_text("cowboys.pdf")
```

Check to see one of the pages.
```{r}
cowboys34 <- cowboys_text[34]

cowboys34
```

Change it to a tidy dataframe.
```{r}
cowboy_tidy <- data.frame(cowboys_text) %>% 
  mutate(text_full = str_split(cowboys_text, pattern = '\\n')) %>% 
  unnest(text_full) %>% 
  mutate(text_full = str_trim(text_full))
```

Take out all the introductory pages that don't have cowboy songs.
```{r}
cowboy_df <- cowboy_tidy %>% 
  slice(-(1:370))
```

Give a new row for each word. 
```{r}
cowboy_tokens <- cowboy_df %>% 
  unnest_tokens(word, text_full) %>% 
  dplyr::select(-cowboys_text)
```

Count all the words up and eliminate all the numbers in the beginning. 
```{r}
cowboy_wordcount <- cowboy_tokens %>% 
  count(word) %>% 
  slice(-(1:155))
```

Time to remove stop words. 
```{r}
cowboy_nonstop_words <- cowboy_tokens %>% 
  anti_join(stop_words)
```

Recount them and get rid of the page numbers
```{r}
nonstop_counts <- cowboy_nonstop_words %>% 
  count(word) %>% 
  slice(-(1:154))
```

Let's look at the top 100 words used in *Songs of the Cattle Trail and Cow Camp*.
```{r}
top_100_words <- nonstop_counts %>% 
  arrange(-n) %>% 
  slice(1:100) 
```

Some cowboy specific stop words snuck through to this step so we're going to manually eliminate them. They are: til, em, fer, yer, git, o'er, yo, thar, and twas. Leaves us with the 91 most used non-stop words.

```{r}
top_words <- top_100_words %>% 
  slice(-c(4,5,6,16,43,45,49,53,88))
```

Let's make a word cloud for *Songs of the Cattle Trail and Cow Camp*.
```{r}
cowboy_cloud <- ggplot(data = top_words, aes(label = word)) +
  geom_text_wordcloud(aes(color = n, size = n), shape = "circle") +
  scale_size_area(max_size = 8) +
  scale_color_gradientn(colors = c("brown","yellow","red")) +
  theme_minimal()

cowboy_cloud
```

- Trail is the most frequently used word in the book
- Surprisingly there are no yeehaws

Time to do sentiment analysis on frequently used cowboy words
```{r}
cowboy_afinn <- top_words %>% 
  inner_join(get_sentiments("afinn"))
```

```{r}

# Plot them: 
ggplot(data = cowboy_afinn, aes(x = value, y = n, fill = word)) +
  geom_col() +
  labs(x = "Sentiment Values", y = "Number of Uses", title = "Most frequent sentimental words in Songs of the Cattle Trail and Cowcamp") +
  theme_minimal()

```

- A lot of love in cowboy songs, it's the most frequently used sentimental word
- The most common negative words by far were about death. Cowboys also have to contend with things that broke (or perhaps being broke financially), getting cut, and fighting.

Source Citation: Lomax, Alan. Songs of the Cattle Trail and Cow Camp. Duell, Sloan, and Pearce, 1909. Google Books, source:"http://books.google.com/books?id=XqnNYZW9olsC&amp;oe=UTF-8". 
